Feeding Dynamic Behavior into Optimization: Simulation-based Multi-objective Overtime Planning

### 2014-09-03 data/result/Article/Latex/Article.tex
Mudanças no texto do artigo

\documentclass[conference]{IEEEtran}
\usepackage{graphicx}
\usepackage{array}
\usepackage{bigstrut}
\usepackage{rotating}
\usepackage{multirow}
\usepackage{booktabs}
\hyphenpenalty=10000

@@

\maketitle
\begin{abstract}
IT professionals are particularly subjected to working overtime due to difficulties in estimation, requirement specification, and measuring progress during a software development life-cycle. However, excessive overtime has detrimental effects on developers and the software they produce. This contrast creates the need for overtime-planning strategies to support project managers in taking the most from as few overtime hours as possible. In this paper we introduce a formulation for the overtime-planning problem which extends the state-of-art by taking into consideration both the positive effects of overtime on productivity and its negative effects on product quality. We use heuristic optimization to explore close to optimal overtime allocations under this formulation. We present an empirical study that compared our approach with current industrial practices and a similar formulation without negative effects. Results show that suppressing the flip-side of the gains brought by overtime may lead to wrong decisions. For instance, excessive allocation of overtime in large projects may lead a manager to underestimate project cost and makespan by 5\% and 3\%, respectively, due to longer testing activities required to fix the exceeding errors introduced by overtime. Evidence also supports an overtime-management strategy used in the industry that concentrates overtime in the second half of a project schedule to compensate for early delays - about 60\% of the overtime allocations produced under this strategy are optimal.
\end{abstract}

@@

Despite the recent growth in works on search-based approaches to software project management \cite{ferrucci:2014}, there is a single previous study that addresses techniques to handle overtime planning \cite{Ferrucci:2013}. The approach uses a multi-objective genetic algorithm to identify the activities providing the best leverage in terms of reducing project makespan and risk of schedule overrun for an amount of overtime work. However, the former paper does not take account of the negative impact of overtime and the consequent rise in errors produced by tired developers. Previous studies \cite{demarco:1982}, \cite{akula:2008} have presented empirical evidence that this is an important factor: the work required to fix the exceeding errors may subdue the gains in schedule brought by working overtime, as well as increase project cost more than expected by stakeholders. The results presented in this paper confirm that ignoring the loss in work quality may lead project managers to incorrect decisions. Furthermore, the present paper finds empirical support for a practice currently used by the industry for overtime allocation.

This paper extends the state-of-the-art by taking into account the documented impact of overtime on the quality of the services provided by developers working in extended hours. It introduces a new formulation for the overtime planning problem which combines search-based optimization with a scenario-based simulation that takes into account an increase in the number of errors injected into a software product by a team working overtime. We compared our results to current practices on regard of overtime management and provide insights to support project managers on deciding when to use overtime work. By doing this, we found scientific support for an overtime-management practice used in the industry, that is, evidence aligned with project management intuition. Our results show that, on average, 60\% of the allocations that might be built according to this strategy are optimal. We also compared our results with a similar model that ignores the dynamics of error generation caused by overtime,  showing that wrong decisions may be taken if the increase in error generation rates is not taken into account. For instance, excessive allocation of overtime in large projects may lead a manager to underestimate project cost and makespan, respectively, by 5\% and 3\% due to longer testing activities required to fix the exceeding errors introduced by overtime. The primary contributions of this paper are:

@@

We depart from a set of work packages to be analyzed, designed, coded, and tested. A project is represented as an acyclic directed graph $<$WP, WPDEP$>$ consisting of a node set WP=\{$wp_1, wp_2, ..., wp_n$\} of work packages and an edge set WPDEP=\{($wp_i, wp_j$): i$\not=$j, 1$\leq$i$\leq$n, 1$\leq$j$\leq$n\} of precedence dependencies amongst work packages. Each $wp_i$ is described by the amount of effort required to elicit its requirements, the amount of effort to design it, code it, and test it as if these tasks were performed by a mean developer in a regular work environment. A work package is also described by the number of errors its requirements analysis, design, and coding activities might generate in a similar context. On Section II.B we discuss how these parameters were filled for our instances. Each dependence in WPDEP represents a work package ($wp_i$) whose development can only start after the requirements for a second work package ($wp_j$) are known.

From these work packages, we develop a project schedule. Our schedule is a partial order of software development activities on which four types of activities are created for each work package: requirement analysis, design, coding, and testing. A testing activity is always preceded by a coding activity, which is preceded by a design activity, which comes after a requirement analysis activity. These dependencies represent the basic workflow of modeling the problem, designing the solution, implementing its source-code, and testing its features. The schedule also includes dependencies amongst work packages: the requirements analysis activity for work package $wp_i$ cannot start before the requirements analysis activity for $wp_j$ is concluded for all $(wp_i, wp_j)$ pairs in WPDEP.

A schedule is an acyclic directed graph $<$AC, ACDEP$>$ consisting of a node set AC=\{$a_1, a_2, ..., a_n$\} of activities and an edge set ACDEP=\{($a_i, a_j$): i$\not=$j,1$\leq$i$\leq$m,1$\leq$j$\leq$m\} of finish-start (FS) dependencies amongst activities. A FS dependency A$\rightarrow$B indicates that work on activity B can only start when activity A is finished. Each activity is classified as either requirement analysis, design, coding, or testing. Analysis, design, and coding activities are described by the expected effort required to produce their results and the expected number of errors they might generate if performed by a mean developer in a regular environment. Testing activities are described by the average effort required to identify and correct an error, which is collected from information data in work packages.

@@

The $<$ACT, ACDEP$>$ graph is comprised of a set of paths $\Pi$ denoting the precedence-respecting sequences of activities that must be executed to complete the project. Makespan is given by the longest path in $\Pi$ (critical path) and is calculated by Equation (1). The duration of an activity is calculated by a simulator according to the characteristics of the activity and its execution setup (including the number of daily overtime hours assigned to it). The simulator calculates the number of errors produced by each development activity and propagates these errors to succeeding activities. For testing activities, the simulator calculates the effort required to identify and correct all errors received from former activities.

@@

The last objective, project cost (CST), is the sum of each activity’s cost. The cost of an activity depends on the number of regular work hours and overtime hours its developers consume to produce its expected results and is related to the duration calculated by the simulator. Again, we modeled the problem according to Brazilian laws: if a regular working hour costs X, each of the first two overtime hours costs 120\%.X and the next two overtime hours cost 150\%.X each.

@@

Time is managed in a continuous fashion. Simulation starts at time zero and is comprised of consecutive {\it simulation steps}, each advancing the simulator's clock by an infinitesimal increment. At each step, all alive simulation objects are allowed to update their internal state. Simulation objects are Java objects that implement an interface defining a protocol through which they are controlled by the simulator. Their state is defined by the relevant characteristics of the real-world element they represent in the simulation. An object's state is kept in its instance variables. State adjustment operations are coded in methods activated by the simulator: {\it init()}, to initialize the object at time zero; {\it start()}, called when all dependencies for an object are finished; {\it step()}, called at each simulation step; and {\it finish()}, called when the object signals its conclusion. The behavior of an object is represented by the history of changes observed in its internal state during the simulation.

@@

Activities are separated into development and testing activities. Development activities are described by an expected amount of work to be performed (in developer-hours) and an amount of errors expected to be generated while this work is performed. The duration of a given activity depends on the amount of work to be performed and the productivity of the developer assigned to it. We calibrated the model using data from Jones \cite{Jones:2000}. Given a work package measured in function-points (FP), Jones \cite{Jones:2000} presents average distributions of effort required for each development phase. For instance, for projects of about 1,000 function-points, requirement analysis activities consume 7\% of project effort, design activities consume about 10\%, coding activities consume about 30\%, and testing consumes 30\%. The remaining effort is related to management, documentation, and other activities. In our model, this effort is proportionally distributed among the analysis, design, coding, and testing activities. Given this distribution, the size of a work package, and an average productivity of 27.8 function-points per developer-month \cite{Jones:2000}, we calculate the expected duration for each activity.

Error dynamics is trickier. When an activity starts, it collects all errors produced by its former activities (if any) carrying them toward the end product. Abdel-Hamid and Madnick \cite{Abdel-Hamid:1991} present a model for error reproduction which states that errors inherited by an activity are not only forwarded to succeeding activities, but they contribute to the generation of new errors. Jones \cite{Jones:2000} suggests that the average number of defects introduced by each kind of activity is 1 error/FP in requirement analysis, 1.25 errors/FP in design, and 1.75 errors/FP in coding activities. Based on these data, our model assumes that each activity introduces 1 new error/FP, design activities regenerate 25\% of errors it receives from analysis, and coding regenerates 33\% of the errors received from design. These regeneration rates lead to the mark of 4.0 errors/FP on average after the implementation and are compatible with the former model \cite{Abdel-Hamid:1991} and source \cite{Jones:2000}.

Testing activities receive the results of development activities and aim at identifying and correcting the errors introduced in the product. These activities are characterized by the average effort required to identify and correct an error. Their duration depends on this parameter along with the assigned developer’s productivity. Given the amount of work expected to be invested in the testing phase \cite{Jones:2000} (i.e., 30\% of total effort for 1,000 FP projects) and the expected number of errors reaching this activity in a regular situation (4 errors/FP), we calculate the average time required to identify and correct each error and use this value on testing activities.

@@

The scenario used in our simulation describes the dynamics of working overtime and its influence on developer's productivity and error generation rates. If a developer is expected to work more than 8 hours per day in a given activity, his productivity will be increased proportionally to the extra number of work hours, reaching up to a 50\% increase if working to the upper limit of 12 hours per day. However, according to deMarco's law \cite{demarco:1999}, a developer working more than the regular daily hours may produce software below the quality standard. The more hours a developer works, the more tired he becomes. Tired developers become less careful to the work they are performing and, consequently, they tend to introduce more errors in the software. We have used data from \cite{Abdel-Hamid:1991}  to estimate the increase in the number of errors produced according to the number of hours worked by a developer. Such data suggest that software quality tends to deteriorate if developers work more than 9 hours/day and that a developer working 12 hours/day commits 50\% more errors than while working on a regular shift.

@@

\noindent
\textbf{Representation}. A candidate solution for OPP is a sequence of m integer numbers, m being the number of activities. Each number in the solution pertains to the [0, 8] integer interval and represents a multiple of 30-minutes overtime for a given activity. Thus, if the number 4 is assigned to activity A, A will receive 1.5 hours of overtime work. Each individual solution being evolved by NSGAII is evaluated through simulation. The simulator activates the overtime scenario upon each activity in the schedule according to the number of overtime hours assigned to that activity in the solution. Then, the simulator calculates the number of overtime hours, project makespan, and project cost for that solution.

\noindent
\textbf{Crossover}. We reused the crossover operator proposed by Ferrucci et al. \cite{Ferrucci:2013}. This operator was adopted because it was shown to outperform the one-point crossover operator commonly used in NSGAII for their formulation for the overtime planning problem. Instead of exchanging genetic material of two parents P$_{1}$ and P$_{2}$ after a point of cut C $\in$ [1..m], the resulting offspring O$_{1}$ and O$_{2}$ produced by this operator are defined by the equations below (given a random variable p with uniform distribution). The operator is executed with 50\% probability; otherwise, the parents are copied to the next generation. Parents are selected by binary tournament.

@@

\noindent
\textbf{RQ1 (Validation):} How does NSGAII perform if compared to random search on finding solutions for OPP? This is a ``sanity check" to determine if our formulation allows a systematic search to outperform random search. If such cannot be verified, either the formulation is poorly designed, parameters are not adequate, or the problem is simple enough to allow a non-systematic procedure to find good solution.

\noindent
\textbf{RQ2 (Competitiveness):} How does NSGAII perform if compared to currently used overtime planning approaches? Besides beating random search, the proposed approach has to outperform current management practices related to overtime planning. In this sense, we compare the results produced by NSGAII with overtime planning practices used in the industry.

\noindent
\textbf{RQ3 (Usefulness):} Do results of overtime planning change if we consider the dynamics of error generation and the loss of quality caused by overtime? To address this issue, we compare the results produced by NSGAII with and without considering an increase in error generation due to overtime. Different results denote that a manager may pick a distinct overtime planning strategy if error dynamics are not considered.

@@

Table I presents information about the selected instances for the experimental analysis. The {\it TF} and {\it DF} columns count, respectively, the number of transaction and data functions for each instance. These values are related to the number of function points of the instance, presented in the last column ({\it FP}). The {\it WP} column presents the number of groups on which the data and transaction functions were distributed to form the project schedule, the {\it ACT} column shows the number of activities in the schedule and the {\it DEP} column shows the number of dependencies among these activities.

% Table generated by Excel2LaTeX from sheet 'tabela 1'
\begin{table}[htbp]
  \centering
  \caption{Instances used in the experimental studies}
    \begin{tabular}
		{lcccccc
%			p{0.7cm} 
%			>{\centering\arraybackslash}p{1.5cm} 
%			>{\centering\arraybackslash}p{1.3cm} 
%			>{\centering\arraybackslash}p{1.3cm} 
%			>{\centering\arraybackslash}p{1.3cm}
		}
    \toprule
		Name & TF & DF & WP & ACT & DEP & FP \\
    \midrule
    ACAD  & 39    & 7     & 10    &  40  &  39  & 185 \\
    WMET  & 48    & 7     & 11    &  44  &  33  & 225 \\
    WAMS  & 67    & 8     & 15    &  60  &  45  & 381 \\
    PSOA  & 65    & 9     & 18    &  72  &  84  & 290 \\
    OMET  & 129   & 22    & 21    &  84  &  63  & 635 \\
    PARM  & 98    & 21    & 27    & 108  &  91  & 451 \\
    \bottomrule
    \end{tabular}%
\end{table}%

ACAD is an academic administrative system that allows managing classes, students, registrations, and teachers in an university setting. WMET allows users to add meteorological observations to a database and provides weather-related reports. WAMS is a message routing system responsible for receiving and routing air traffic control messages. PSOA is a personnel management project used by a large company for authentication, authorization, and centralization of references from enterprise systems. OMET is a system which manages, stores, and delivers meteorological information, including messages and statistical information regarding the weather. PARM stores configuration and setting values used by other systems, allowing for fast configuration of user profiles and sharing configurations amongst distinct applications.

@@

To compare two Pareto fronts F$_{A}$ and F$_{B}$ we need surrogate measures of their quality. These measures are called {\it quality indicators}. A quality indicator receives a set of points defined in the objective space and produces a real number. The result for a given front F$_{A}$ can be compared to the result for a second front F$_{B}$, leading to conclusions about the quality of these fronts and, consequently, the algorithms and configurations that produced them. Quality indicators usually compare each front with a 'reference front' F$_{R}$ built from the non-dominated solutions of all Pareto fronts under comparison. The ideal reference front would be the optimal set of non-dominated solutions for a given instance, but these are rarely available for practical purposes. Thus, the union of all available fronts followed by the elimination of duplicate and dominated solutions is used as a proxy to the ideal reference front.

Given F$_{R}$ and two fronts to be compared a number of quality indicators can be defined. They usually address two properties of Pareto fronts: convergence and diversity. Convergence indicators measure how close a given front F$_{A}$ is to the reference front. Diversity indicators measure the amount of the solution space that is covered by a given front F$_{A}$. For the purpose of our study, we selected three indicators: Contributions (I$_{C}$), Hypervolume (I$_{HV}$), and Generational Distance $(I_{GD})$. These indicators are measured in the [0, 1] interval. I$_{C}$ is a convergence indicator that represents the proportion of solutions in a given front F$_{A}$ that also pertain to the reference front. Good Pareto fronts have high I$_{C}$, thus contributing heavily to the composition of F$_{R}$. I$_{HV}$ calculates the volume of the objective space that is covered by solutions of a given Pareto front \cite{Durillo:2009}. It mixes convergence to the reference front and diversity amongst the selected solutions comprising a given front. High-quality Pareto fronts present high hypervolume, thus covering a large extension of the objective space. Finally,  is a convergence indicator that measures the distance between a front and the reference front \cite{vanveldhuizen:1998}. Good fronts are closer to the reference front and thus present low generational distance.

\subsection{Parameter Setting for the Algorithm}
Section III presented the selected configuration for most components of NSGAII in order to fine-tune the algorithm to the OPP. However, two components were not addressed in that section because they required experimental analysis to be defined: the stop criteria, expressed in terms of a maximum number of fitness function evaluations NSGAII is allowed to perform while searching for solutions, and population size, that is, the size of the genetic pool evolved by NSGAII as a multiplier to the number of activities in a given instance.

>>>>

\documentclass[conference]{IEEEtran}
\usepackage{graphicx}
\usepackage{array}
\usepackage{bigstrut}
\usepackage{rotating}
\usepackage{multirow}
\usepackage{booktabs}
\usepackage{soul}

\hyphenpenalty=10000

@@

\maketitle
\begin{abstract}
IT professionals are particularly subjected to working overtime due to difficulties in estimation, requirement specification, and measuring progress during a software development life-cycle. However, excessive overtime has detrimental effects on developers and the software they produce. This contrast creates the need for overtime-planning strategies to support project managers in taking the most from as few overtime hours as possible. In this paper we introduce a formulation for the overtime-planning problem which extends the state-of-art by taking into consideration both the positive effects of overtime on productivity and its negative effects on product quality. We use heuristic optimization to explore close to optimal overtime allocations under this formulation. We present an empirical study that compared our approach with current industrial practices and a similar formulation without negative effects. Results show that suppressing the flip-side of the gains brought by overtime may lead to wrong decisions. For instance, excessive allocation of overtime in large projects may lead a manager to underestimate project cost and makespan \hl{by 5\% and 3\%, respectively}, due to longer testing activities required to fix the exceeding errors introduced by overtime. Evidence also supports an overtime-management strategy used in the industry that concentrates overtime in the second half of a project schedule to compensate for early delays - \hl{about 60\% of the overtime allocations produced under this strategy are optimal}.
\end{abstract}

@@

Despite the recent growth in works on search-based approaches to software project management \cite{ferrucci:2014}, there is a single previous study that addresses techniques to handle overtime planning \cite{Ferrucci:2013}. The approach uses a multi-objective genetic algorithm to identify the activities providing the best leverage in terms of reducing project makespan and risk of schedule overrun for an amount of overtime work. However, the former paper does not take account of the negative impact of overtime and the consequent rise in errors produced by tired developers. Previous studies \cite{demarco:1982} \cite{akula:2008} have presented empirical evidence that this is an important factor: the work required to fix the exceeding errors may subdue the gains in schedule brought by working overtime, as well as increase project cost more than expected by stakeholders.

This paper extends the state-of-the-art by taking into account the documented impact of overtime on the quality of the services provided by developers working in extended hours. It introduces a new formulation for the overtime planning problem which combines search-based optimization with a scenario-based simulation that takes into account an increase in the number of errors injected into a software product by a team working overtime. We compared our results to current overtime management practices and provide insights to support project managers on deciding when to use overtime work. By doing this, we found empirical support for an overtime allocation practice used in the industry, that is, scientific evidence aligned with project management intuition. \hl{Our results show that, on average, 60\% of the allocations that might be built according to this strategy are optimal.} We also compared our results with a similar model that ignores the dynamics of error generation caused by overtime,  showing that project managers may take wrong decisions if the increase in error generation rates is not taken into account. For instance, \hl{excessive allocation of overtime in large projects may lead a manager to underestimate project cost and makespan, respectively, by 5\% and 3\% due to longer testing activities required to fix the exceeding errors introduced by overtime.} The primary contributions of this paper are:

@@

We depart from a set of work packages to be analyzed, designed, coded, and tested. A project is represented as an acyclic directed graph $<$WP, WPDEP$>$ consisting of a node set WP=\{$wp_1, wp_2, ..., wp_n$\} of work packages and an edge set WPDEP=\{($wp_i, wp_j$): i$\not=$j, 1$\leq$i$\leq$n, 1$\leq$j$\leq$n\} of precedence dependencies amongst work packages. Each $wp_i$ is described by the amount of effort required to elicit its requirements, the amount of effort to design it, code it, and test it as if these tasks were performed by a mean developer in a regular work environment. A work package is also described by the number of errors its requirements analysis, design, and coding activities might generate in a similar context. On Section II.B we discuss how these parameters were filled for our instances. Each dependence in WPDEP represents a work package ($wp_j$) whose development can only start after the requirements for a second work package ($wp_i$) are known.

From these work packages, we develop a project schedule. A schedule is a partial order of four types of software development activities created for each work package: requirement analysis, design, coding, and testing. A testing activity is always preceded by a coding activity, which is preceded by a design activity, which comes next to a requirement analysis activity. These dependencies represent the workflow of modeling the problem, designing the solution, implementing its source-code, and testing its features. The schedule also includes dependencies amongst work packages: the requirements analysis activity for work package $wp_j$ cannot start before the requirements analysis activity for $wp_i$ is concluded for all $(wp_i, wp_j)$ pairs in WPDEP.

A schedule is an acyclic directed graph $<$AC, ACDEP$>$ consisting of a node set AC=\{$a_1, a_2, ..., a_m$\} of activities and an edge set ACDEP=\{($a_i, a_j$): i$\not=$j,1$\leq$i$\leq$m,1$\leq$j$\leq$m\} of finish-start (FS) dependencies amongst activities. A FS dependency A$\rightarrow$B indicates that work on activity B can only start when activity A is finished. Each activity is classified as either requirement analysis, design, coding, or testing. Analysis, design, and coding activities are described by the expected effort required to produce their results and the expected number of errors they might generate if performed by a mean developer in a regular work environment. Testing activities are described by the average effort required to identify and correct an error, which is collected from information provided in work packages.

@@

The $<$AC, ACDEP$>$ graph is comprised of a set of paths $\Pi$ denoting the precedence-respecting sequences of activities that must be executed to complete the project. Makespan is given by the longest path in $\Pi$ (critical path) and is calculated by Equation (1). The duration of an activity is calculated by a simulator according to the characteristics of the activity and its execution setup (including the number of daily overtime hours assigned to it). The simulator calculates the number of errors produced by each development activity and propagates these errors to succeeding activities. For testing activities, the simulator calculates the effort required to identify and correct all errors received from former activities.

@@

The last objective, project cost (CST), is the sum of each activity's cost. The cost of an activity depends on the number of regular work hours and overtime hours its developers consume to produce its expected results and is related to the duration calculated by the simulator. Again, we modeled the problem according to Brazilian laws: if a regular working hour costs X, each of the first two overtime hours costs 120\%.X and the next two overtime hours cost 150\%.X each.

@@

Time is managed in a continuous fashion. Simulation starts at time zero and is comprised of consecutive {\it simulation steps}, each advancing the simulator's clock by an infinitesimal increment. At each step, all alive simulation objects are allowed to update their internal state. Simulation objects are Java objects that implement an interface defining a protocol through which they are controlled by the simulator. Their state is defined by the relevant characteristics of the real-world element they represent in the simulation and is kept in instance variables. State adjustment operations are coded in methods activated by the simulator: {\it init()}, to initialize the object at time zero; {\it start()}, called when all dependencies for an object are finished; {\it step()}, called at each simulation step; and {\it finish()}, called when the object signals its conclusion. The behavior of an object is represented by the history of changes observed in its internal state during the simulation.

@@

Activities are separated into development and testing. Development activities are described by an expected amount of work to be performed (in developer-hours) and an amount of errors expected to be generated while this work is performed. The duration of a given activity depends on the amount of work to be performed and the productivity of the developer assigned to it. We calibrated the model using data from Jones \cite{Jones:2000}. Given a work package measured in function-points (FP), Jones \cite{Jones:2000} presents average distributions of effort required for each development phase. For instance, for projects of about 1,000 function-points, requirement analysis activities consume 7\% of the total project effort, design activities consume about 10\%, coding activities consume about 30\%, and testing consumes 30\%. The remaining effort is related to management, documentation, and other activities. In our model, this effort is proportionally distributed among the analysis, design, coding, and testing activities. Given this distribution, the size of a work package, and an average productivity of 27.8 function-points per developer-month \cite{Jones:2000}, we calculate the expected duration for each activity.

Error dynamics is trickier. When an activity starts, it collects all errors produced by its former activities (if any) carrying them toward the end product. Abdel-Hamid and Madnick \cite{Abdel-Hamid:1991} present a model for error reproduction which states that errors inherited by an activity are not only forwarded to succeeding activities, but they contribute to the generation of new errors. Jones \cite{Jones:2000} suggests that the average number of defects introduced by each kind of activity is 1 error/FP in requirement analysis, 1.25 errors/FP in design, and 1.75 errors/FP in coding activities. Based on these data, our model assumes that each activity introduces 1 new error/FP, design activities regenerate 25\% of the errors it receives from analysis, and coding regenerates 33\% of the errors received from design. These regeneration rates lead to the mark of 4 errors/FP on average after the implementation and are compatible with the former model \cite{Abdel-Hamid:1991} and source \cite{Jones:2000}.

Testing activities receive the results of development activities and aim at identifying and correcting the errors introduced in the product. These activities are characterized by the average effort required to identify and correct an error. The duration of testing activities depends on this parameter along with the assigned developers productivity. Given the amount of work expected to be invested in the testing phase \cite{Jones:2000} (i.e., 30\% of total effort for 1,000 FP projects) and the expected number of errors reaching this activity in a regular situation (4 errors/FP), we calculate the average time required to identify and correct an error and use this value to estimate the duration of testing activities.

@@

The scenario used in our simulation describes the dynamics of working overtime and its influence on developer's productivity and error generation rates. If a developer is expected to work more than 8 hours per day in a given activity, his productivity will be increased proportionally to the extra number of work hours, reaching up to a 50\% increase if working to the upper limit of 12 hours per day. However, according to deMarco's law \cite{demarco:1999}, a developer working more than the regular daily hours may produce software below the quality standard. The more hours a developer works, the more tired he becomes. Tired developers become less careful to the work they are performing and, consequently, they tend to introduce more errors in the software. We have used data from \cite{Abdel-Hamid:1991}  to estimate the increase in the number of errors produced according to the number of hours worked by a developer. Such data suggest that software quality tends to deteriorate if developers work more than 9 hours/day and that a developer working 12 hours/day commits 50\% more errors than developers working on a regular shift.

@@

\vspace{3px}
\noindent
\textbf{Representation}. A candidate solution for OPP is a sequence of $m$ integer numbers, $m$ being the number of activities. Each number in the solution pertains to the [0, 8] integer interval and represents a multiple of 30-minutes overtime for a given activity. Thus, if the number 4 is assigned to activity A, A will receive 1.5 hours of overtime work. Each individual solution being evolved by NSGAII is evaluated through simulation. The simulator activates the overtime scenario upon each activity in the schedule according to the number of overtime hours assigned to that activity in the solution. Then, the simulator calculates the number of overtime hours, project makespan, and project cost for that solution.

\vspace{3px}
\noindent
\textbf{Crossover}. We reused the crossover operator proposed by Ferrucci et al. \cite{Ferrucci:2013}. This operator was adopted because it was shown to outperform the one-point crossover operator commonly used in NSGAII for their formulation of the overtime planning problem. Instead of exchanging genetic material of two parents P$_{1}$ and P$_{2}$ after a point of cut C $\in$ [1..$m$], the resulting offspring O$_{1}$ and O$_{2}$ produced by this operator are defined by the equations below (given a random variable p with uniform distribution). The operator is executed with 50\% probability; otherwise, the parents are copied to the next generation. Parents are selected by binary tournament.

@@

\vspace{3px}
\noindent
\textbf{RQ1 (Validation):} How does NSGAII perform if compared to random search on finding solutions for OPP? This is a ``sanity check" to determine if our formulation allows a systematic search to outperform random search. If such cannot be verified, either the formulation is poorly designed, parameters are not adequate, or the problem is simple enough to allow a non-systematic procedure to find good solutions.

\vspace{3px}
\noindent
\textbf{RQ2 (Competitiveness):} How does NSGAII perform if compared to currently used overtime planning approaches? Besides beating random search, the proposed approach has to outperform current management practice related to overtime planning. In this sense, we compare the results produced by NSGAII with overtime planning practices used in the industry.

\vspace{3px}
\noindent
\textbf{RQ3 (Usefulness):} Do results of overtime planning change if we consider the dynamics of error generation and the loss of quality caused by overtime? To address this issue, we compare the results produced by NSGAII with and without an increase in error generation due to overtime. Different results denote that a manager may pick a distinct overtime planning strategy if error dynamics are not considered.

@@

Table I presents information about the selected instances used in the experimental analysis. The {\it TF} and {\it DF} columns count, respectively, the number of transaction and data functions for each instance. These values are related to the number of function points of the instance, presented in the last column ({\it FP}). The {\it WP} column presents the number of groups on which the data and transaction functions were distributed to form the project schedule, the {\it ACT} column shows the number of activities in the schedule, and the {\it DEP} column shows the number of dependencies among these activities.

% Table generated by Excel2LaTeX from sheet 'tabela 1'
\begin{table}[htbp]
  \centering
  \caption{Instances used in the experimental studies}
    \begin{tabular}
		{lcccccc}
    \toprule
		Name & TF & DF & WP & ACT & DEP & FP \\
    \midrule
    ACAD  & 39    & 7     & 10    &  40  &  39  & 185 \\
    WMET  & 48    & 7     & 11    &  44  &  33  & 225 \\
    WAMS  & 67    & 8     & 15    &  60  &  45  & 381 \\
    PSOA  & 65    & 9     & 18    &  72  &  84  & 290 \\
    OMET  & 129   & 22    & 21    &  84  &  63  & 635 \\
    PARM  & 98    & 21    & 27    & 108  &  91  & 451 \\
    \bottomrule
    \end{tabular}%
\end{table}%

ACAD is an academic administrative system that allows managing classes, students, registrations, and teachers in an university setting. WMET allows users to add meteorological observations to a database and provides weather-related reports. WAMS is a message routing system responsible for receiving and routing air traffic control messages. PSOA is a personnel management project used by a large company for authentication, authorization, and centralization of references from enterprise systems. OMET manages, stores, and delivers meteorological information, including messages and statistical data about the weather. PARM stores settings values used by many systems, allowing for fast configuration of user profiles and sharing configurations amongst distinct applications.

@@

To compare two Pareto fronts F$_{A}$ and F$_{B}$ we need surrogate measures of their quality. These measures are called {\it quality indicators}. A quality indicator receives a set of points defined in the objective space and produces a real number. The result for a given front F$_{A}$ can be compared to the result for a second front F$_{B}$, leading to conclusions about the quality of these fronts and, consequently, the algorithms and configurations that produced them. Quality indicators usually compare each front with a `reference front' F$_{R}$ built from the non-dominated solutions of all Pareto fronts under comparison. The ideal reference front would be the optimal set of non-dominated solutions for a given instance, but these are rarely available for practical purposes. Thus, the union of all available fronts followed by the elimination of duplicate and dominated solutions is used as a proxy to the ideal reference front.

Given F$_{R}$ and two fronts to be compared a number of quality indicators can be defined. They usually address two properties of Pareto fronts: convergence and diversity. Convergence indicators measure how close a given front F$_{A}$ is to the reference front. Diversity indicators measure the amount of solution space that is covered by a given front F$_{A}$. For the purpose of our study, we selected three indicators: Contributions (I$_{C}$), Hypervolume (I$_{HV}$), and Generational Distance $(I_{GD})$. These indicators are measured in the [0, 1] interval. I$_{C}$ is a convergence indicator that represents the proportion of solutions in a given front F$_{A}$ that also pertain to the reference front. Good Pareto fronts have high I$_{C}$, thus contributing heavily to the composition of F$_{R}$. I$_{HV}$ calculates the volume of the objective space that is covered by solutions of a given Pareto front \cite{Durillo:2009}. It mixes convergence to the reference front and diversity amongst the solutions comprising a given front. High-quality Pareto fronts present high $I_{HV}$, thus covering a large extension of the objective space. Finally, I$_{GD}$ is a convergence indicator that measures the distance between a front and the reference front \cite{vanveldhuizen:1998}. Good fronts are closer to the reference front and thus present low $I_{GD}$.

\subsection{Parameter Setting for the Algorithm}
Section III presented the selected configuration for most components of NSGAII in order to fine-tune the algorithm to the OPP. However, two components were not addressed in that section because they required experimental analysis to be defined: the stop criteria, expressed in terms of a maximum number of fitness function evaluations NSGAII is allowed to perform while searching for solutions, and population size, the size of the genetic pool evolved by NSGAII expressed as a multiplier to the number of activities in a given instance.
